---
title: "3 - exercise"
format: 
  html: 
    embed-resources: true
editor: visual
---

You're going to practice the same charting pipeline we built in the lesson, but on a stock of your choice. Pick something you have some opinion about - a company you buy from, a bank you use, an airline you love to hate. The point is to see that this whole workflow generalizes.

### 1. Load libraries that you'll need

Load the following libraries: `tidyverse`, `tidyquant`, `quantmod`, `plotly`, and `TTR`.

```{r}
#| label: load-libraries
#| warning: false
#| message: false

# load libraries

```

### 2. Download stock price data

Take a look back to the code we used for the first lesson and exercise. Create a tibble called `prices_yourtickerhere`. For example, if you choose to examine Walmart, your prices tibble would be `prices_WMT`. Download one year of stock prices for the stock of your choice.

```{r}
#| label: get-stock-data

```

### 3. Create a simple line chart of your stock prices

Chart your stock prices. We've done this plenty of times now. Be sure to use the adjusted close price (since that is adjusted for splits and dividends). Title your plot "Your stock here Adjusted Closing Prices." Label your axes.

```{r}
#| label: simple-line-chart

```

### 4. Add moving averages

Add a 30 day and a 60 day moving average to your chart. Be sure to add a subtitle to explain what those extra lines are.

```{r}
#| label: sma-chart
#| warning: false

# add 30 and 60 day moving averages

```

### 5. Bollinger bands

Drop the 60 day moving average and add Bollinger Bands. Add a meaningful title so people know what's going on. Use three standard deviations for the width of the "trench" in the bands.

```{r}
#| label: bollinger-bands
#| warning: false

# Bollinger Bands, 30 day SMA

```
